---
title: "MHC Class I Epitope Analysis Pipeline"
author: "Ibnu Ariyanto"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: show
  pdf_document:
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)
```
## Overview

This report implements a comprehensive MHC Class I epitope analysis pipeline based on NetMHCpan output. It standardizes inputs, assigns binding classes, scores promiscuity across HLA alleles, filters candidates using simple physicochemical features, and exports publication-ready figures and tables.

## 1. Setup & Data Import

### 1.1 Packages

```{r packages, echo=FALSE}
pkgs <- c(
  "tidyverse", "ggplot2", "ggrepel", "pheatmap", "RColorBrewer",
  "viridis", "gridExtra", "knitr", "kableExtra", "writexl",
  "VennDiagram", "UpSetR", "corrplot", "scales", "patchwork"
)
install.packages(setdiff(pkgs, rownames(installed.packages())))

library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(gridExtra)
library(knitr)
library(kableExtra)
library(writexl)
library(VennDiagram)
library(UpSetR)
library(corrplot)
library(scales)
library(patchwork)
```

### 1.2 Input files

Update these paths to match your working directory.
```{r data_import, echo=FALSE}
df_all    <- readr::read_csv("peptide_table_40dba94e.csv")
df_pat    <- readr::read_csv("MHCI_peptide_table_e04f7760_sort.csv")
dist_all  <- readr::read_csv("allele_distances_table_40dba94e.csv")
dist_pat  <- readr::read_csv("MHCI_allele_distances_table_e04f7760_sort.csv")

# Standardize column names

df_all <- df_all %>%
  rename_with(tolower) %>%
  rename(
    peptide = peptide,
    allele = allele,
    el_pct = `netmhcpan_el percentile`,
    el_score = `netmhcpan_el score`,
    median_rank = `median binding percentile`
  )

df_pat <- df_pat %>%
  rename_with(tolower) %>%
  rename(
    peptide = peptide,
    allele = allele,
    el_pct = `netmhcpan_el percentile`,
    el_score = `netmhcpan_el score`,
    median_rank = `median binding percentile`
  )
```

1.3 Quick input checks
```{Quick unput, echo=FALSE}
stopifnot(all(c("peptide","allele","el_pct") %in% names(df_all)))
stopifnot(is.numeric(df_all$el_pct))
```
## 2. Binding Classification

NetMHCpan convention (EL %ile):

- Strong binder (SB): EL %ile ≤ 0.5
- Weak binder (WB): EL %ile ≤ 2.0
- Non-binder (NB): otherwise

