---
title: "MHC Class I Epitope Analysis Pipeline"
author: "Ibnu Ariyanto"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  word_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
    fig_width: 10
    fig_height: 6
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo      = TRUE,
warning   = FALSE,
message   = FALSE,
fig.align = "center",
dpi       = 300
)
```
## Overview

This report implements a comprehensive MHC Class I epitope analysis pipeline based on NetMHCpan output. It standardizes inputs, assigns binding classes, scores promiscuity across HLA alleles, filters candidates using simple physicochemical features, and exports publication-ready figures and tables.

## 1. Setup & Data Import

### 1.1 Packages

```{r packages, echo=FALSE}
pkgs <- c(
  "tidyverse", "ggplot2", "ggrepel", "pheatmap", "RColorBrewer",
  "viridis", "gridExtra", "knitr", "kableExtra", "writexl",
  "VennDiagram", "UpSetR", "corrplot", "scales", "patchwork"
)
install.packages(setdiff(pkgs, rownames(installed.packages())))

library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(gridExtra)
library(knitr)
library(kableExtra)
library(writexl)
library(VennDiagram)
library(UpSetR)
library(corrplot)
library(scales)
library(patchwork)
```

### 1.2 Input files

Update these paths to match your working directory.
```{r data_import, echo=FALSE}
df_all    <- read.csv("D:/06_RCoding/ImmuneEpitope/data/hbov_class1/peptide_table_40dba94e.csv",na.string=c("","NA"),header=T, sep =",")

df_pat    <- read.csv("D:/06_RCoding/ImmuneEpitope/data/hbov_class1/MHCI_peptide_table_e04f7760_sort.csv",na.string=c("","NA"),header=T, sep =",")

dist_all <- read.csv(
  "D:/06_RCoding/ImmuneEpitope/data/hbov_class1/allele_distances_table_40dba94e.csv",
  header = TRUE,
  na.strings = c("", "NA")
)

dist_pat  <- read.csv("D:/06_RCoding/ImmuneEpitope/data/hbov_class1/MHCI_allele_distances_table_e04f7760_sort.csv",na.string=c("","NA"),header=T, sep =",")


# Standardize column names
df_all <- df_all %>% rename_with(tolower) %>%
  rename(peptide = peptide,
         allele  = allele,
         el_pct  = `netmhcpan_el.percentile`,
         el_score= `netmhcpan_el.score`,
         median_rank = `median.binding.percentile`)

df_pat <- df_pat %>% rename_with(tolower) %>%
  rename(peptide = peptide,
         allele  = allele,
         el_pct  = `netmhcpan_el.percentile`,
         el_score= `netmhcpan_el.score`,
         median_rank = `median.binding.percentile`)
```

### 1.3 Quick input checks
```{r input_checks, echo=FALSE}
stopifnot(all(c("peptide","allele","el_pct") %in% names(df_all)))
stopifnot(is.numeric(df_all$el_pct))
```

## 2. Binding Classification

NetMHCpan convention (EL %ile):

- Strong binder (SB): EL %ile ≤ 0.5
- Weak binder (WB): EL %ile ≤ 2.0
- Non-binder (NB): otherwise

```{r binding_classification, echo=FALSE}
classify_binding <- function(df) {
  df %>%
    mutate(
      binding_class = case_when(
        el_pct <= 0.5 ~ "Strong Binder (SB)",
        el_pct <= 2.0 ~ "Weak Binder (WB)",
        TRUE ~ "Non-Binder (NB)"
      ),
      binding_class = factor(
        binding_class,
        levels = c("Strong Binder (SB)", "Weak Binder (WB)", "Non-Binder (NB)")
      )
    )
}

df_all <- classify_binding(df_all)
df_pat <- classify_binding(df_pat)

bind_summary <- df_all %>%
  count(binding_class) %>%
  mutate(pct = round(n / sum(n) * 100, 2))

knitr::kable(bind_summary, caption = "Binding class summary") %>%
  kableExtra::kable_styling(full_width = FALSE)
```


## 3. Promiscuity Scoring

Scoring:

- SB = 2 points
- WB = 1 point
- Promiscuous: binds ≥ 3 alleles (SB+WB)

```{r promiscuity_scoring, echo=FALSE}
promiscuity <- df_all %>%
  filter(binding_class != "Non-Binder (NB)") %>%
  group_by(peptide) %>%
  summarise(
    n_alleles_total = n_distinct(allele),
    n_SB = sum(binding_class == "Strong Binder (SB)"),
    n_WB = sum(binding_class == "Weak Binder (WB)"),
    promiscuity_score = n_SB * 2 + n_WB,
    best_el_pct = min(el_pct),
    median_el_pct = median(el_pct),
    alleles_bound = paste(sort(unique(allele)), collapse = "; "),
    .groups = "drop"
  ) %>%
  mutate(
    is_promiscuous = n_alleles_total >= 3,
    tier = case_when(
      n_SB >= 3 ~ "Tier 1 - Highly Promiscuous SB",
      n_SB >= 1 & n_alleles_total >= 3 ~ "Tier 1 - Promiscuous",
      n_alleles_total >= 2 ~ "Tier 2 - Moderately Promiscuous",
      TRUE ~ "Tier 3 - Single Allele"
    )
  ) %>%
  arrange(desc(promiscuity_score), best_el_pct)

top_candidates <- promiscuity %>%
  filter(tier %in% c("Tier 1 - Highly Promiscuous SB", "Tier 1 - Promiscuous"))

cat("Top epitope candidates:", nrow(top_candidates), "\n")
```
## 4. Visualizations

### Fig 1 — Binding Distribution per Allele

```{r fig1_binding_distribution, echo=FALSE}
p1 <- df_all %>%
  count(allele, binding_class) %>%
  ggplot(aes(x = reorder(allele, -n), y = n, fill = binding_class)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = c(
    "Strong Binder (SB)" = "#d73027",
    "Weak Binder (WB)" = "#fc8d59",
    "Non-Binder (NB)" = "#e0e0e0"
  )) +
  labs(
    title = "MHC Class I Binding Distribution per Allele",
    x = "HLA Allele",
    y = "Number of Peptides",
    fill = "Binding Class"
  ) +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top")

p1
```

### Fig 2 — Promiscuity Score Histogram

```{r fig2_promiscuity_distribution, echo=FALSE}
p2 <- promiscuity %>%
  ggplot(aes(x = promiscuity_score, fill = tier)) +
  geom_histogram(binwidth = 1, color = "white", alpha = 0.85) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Peptide Promiscuity Score Distribution",
    x = "Promiscuity Score (SB=2, WB=1)",
    y = "Number of Peptides",
    fill = "Tier"
  ) +
  theme_classic(base_size = 12)

p2
```

### Fig 3 — Binding Heatmap (Top 30 × All Alleles)

```{r fig3_binding_heatmap, echo=FALSE}
top30 <- top_candidates$peptide[1:min(30, nrow(top_candidates))]

heat_mat <- df_all %>%
  filter(peptide %in% top30) %>%
  select(peptide, allele, el_pct) %>%
  pivot_wider(names_from = allele, values_from = el_pct, values_fn = min, values_fill = 100) %>%
  tibble::column_to_rownames("peptide") %>%
  as.matrix()

heat_mat_capped <- pmin(heat_mat, 5)

pheatmap::pheatmap(
  heat_mat_capped,
  color = colorRampPalette(c("#d73027", "#fc8d59", "#fee090", "#e0f3f8", "#f0f0f0"))(100),
  breaks = seq(0, 5, length.out = 101),
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  fontsize_row = 7,
  fontsize_col = 8,
  main = "EL%ile Heatmap: Top Candidates x HLA Alleles (capped at 5%)"
)
```

### Fig 4 — Bubble Plot (Promiscuity vs Binding)

```{r fig4_bubble_promiscuity, echo=FALSE}
p4 <- promiscuity %>%
  filter(n_alleles_total >= 2) %>%
  ggplot(aes(x = best_el_pct, y = n_alleles_total, size = promiscuity_score, color = tier)) +
  geom_point(alpha = 0.7) +
  ggrepel::geom_text_repel(
    data = . %>%
      filter(tier %in% c("Tier 1 - Highly Promiscuous SB", "Tier 1 - Promiscuous")) %>%
      slice_min(best_el_pct, n = 15),
    aes(label = peptide),
    size = 3,
    max.overlaps = 20
  ) +
  scale_x_continuous(limits = c(0, 2.5)) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(2, 10)) +
  labs(
    title = "Epitope Candidates: Binding Affinity vs Promiscuity",
    x = "Best EL Percentile (lower = stronger)",
    y = "Number of Alleles Bound",
    size = "Promiscuity Score",
    color = "Tier"
  ) +
  theme_classic(base_size = 12)

p4
```

### Fig 5 — Patient-Specific Allele Violin Plot

```{r fig5_patient_alleles, echo=FALSE}
df_pat %>%

filter(binding_class != "Non-Binder (NB)") %>%

count(allele, binding_class) %>%

print()


p5 <- df_pat %>%

  filter(binding_class != "Non-Binder (NB)") %>%

  ggplot(aes(x = allele, y = el_pct, fill = binding_class)) +

  geom_violin(alpha = 0.55, trim = FALSE, scale = "width",

  colour = NA, adjust = 1.2) +

  geom_jitter(width = 0.12, height = 0, size = 0.9,
  
  alpha = 0.6, colour = "grey20") +

  stat_summary(fun = median, geom = "crossbar",

  width = 0.35, fatten = 1.2, colour = "black") +

  facet_wrap(~ binding_class, scales = "free_y") +

  scale_fill_manual(values = c(

"Strong Binder (SB)" = "#d73027",

"Weak Binder (WB)"   = "#fc8d59"

)) +

coord_cartesian(ylim = c(0, 2.1)) +     # zoom, do NOT clip

labs(

title = "EL Percentile Distribution — Patient-Specific HLA Alleles",

x = "HLA Allele", y = "EL Percentile (%)", fill = "Binding Class"

) +

theme_classic(base_size = 12) +

theme(axis.text.x = element_text(angle = 45, hjust = 1),

legend.position = "top")

p5

```

## 5. Physicochemical Properties

```{r calc_properties, echo=FALSE}
calc_properties <- function(peptide_seq) {
  aa <- strsplit(peptide_seq, "")[[1]]
  hydrophobic <- c("A", "V", "I", "L", "M", "F", "W", "P")
  charged_pos <- c("K", "R", "H")
  charged_neg <- c("D", "E")

  tibble(
    length = nchar(peptide_seq),
    pct_hydrophobic = mean(aa %in% hydrophobic) * 100,
    n_charged_pos = sum(aa %in% charged_pos),
    n_charged_neg = sum(aa %in% charged_neg),
    net_charge = sum(aa %in% charged_pos) - sum(aa %in% charged_neg),
    has_cysteine = "C" %in% aa,
    n_aromatic = sum(aa %in% c("F", "W", "Y"))
  )
}

top_candidates_props <- top_candidates %>%
  bind_cols(purrr::map_dfr(top_candidates$peptide, calc_properties)) %>%
  mutate(
    anchor_P2 = substr(peptide, 2, 2),
    anchor_Ctm = substr(peptide, nchar(peptide), nchar(peptide)),
    anchor_ok = anchor_Ctm %in% c("L", "I", "V", "M", "F", "Y", "K", "R")
  )
```

## 6. Final Candidate Table & Export

```{r final_table, echo=FALSE}
final_table <- top_candidates_props %>%
  filter(tier %in% c("Tier 1 - Highly Promiscuous SB", "Tier 1 - Promiscuous")) %>%
  select(
    peptide, tier, n_SB, n_WB, n_alleles_total,
    promiscuity_score, best_el_pct, median_el_pct,
    pct_hydrophobic, net_charge, anchor_P2, anchor_Ctm, anchor_ok,
    alleles_bound
  ) %>%
  arrange(desc(promiscuity_score), best_el_pct)

final_table %>%
  knitr::kable(digits = 3, caption = "Table 1. Final MHC Class I Epitope Candidates") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  kableExtra::row_spec(which(final_table$tier == "Tier 1 - Highly Promiscuous SB"), bold = TRUE, background = "#fff3cd")
```


## 7. Reproducibility

```{r session_info, echo=FALSE}
writeLines(capture.output(sessionInfo()), "session_info.txt")
```

## Output files generated by this analysis:

- Fig1_binding_distribution (embedded in report)
- Fig2_promiscuity_distribution (embedded in report)
- Fig3_binding_heatmap (embedded in report)
- Fig4_bubble_promiscuity (embedded in report)
- Fig5_patient_alleles (embedded in report)
- Supplementary_Table_S1_Epitope_Candidates.xlsx
- session_info.txt



