---
title: "MHC-II Peptide Binding Analysis for Vaccine Epitope Selection"
author: "Ibnu Ariyanto"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
    fig_width: 10
    fig_height: 6
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
echo      = TRUE,
warning   = FALSE,
message   = FALSE,
fig.align = "center",
dpi       = 150
)
```

# Background

This report presents an in-silico MHC-II binding analysis of peptide candidates
for vaccine epitope selection. Binding affinity was predicted using **NetMHCIIpan**
(EL method). Lower EL percentile rank indicates stronger predicted binding:
**<= 2%** = strong binder, **<= 10%** = moderate binder.

---

# Setup

## Load packages

```{r packages}
# ```{r packages}
library(tidyverse)
library(readr)
library(pheatmap)
library(factoextra)
library(ggrepel)
library(ggseqlogo)
library(knitr)
library(kableExtra)
library(cowplot)
# ```
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
# ```{r import}
df_wide   <- read.csv("D:/06_RCoding/ImmuneEpitope/data/hbov_class2/consolidated_peptide_table_de2f9162.csv",na.string=c("","NA"),header=T, sep =",")
df_2allele <- read.csv("D:/06_RCoding/ImmuneEpitope/data/hbov_class2/MHCII_consolidated_peptide_table_5182db8f_sort.csv",na.string=c("","NA"),header=T, sep =",")
df_long   <- read.csv("D:/06_RCoding/ImmuneEpitope/data/hbov_class2/peptide_table_de2f9162.csv",na.string=c("","NA"),header=T, sep =",")


# Safe column references (robust to name variation)
median_col     <- grep("overall",    colnames(df_wide),    value = TRUE)[1]
global_col     <- grep("global",     colnames(df_2allele), value = TRUE)[1]
drb12_col      <- grep("DRB1.12",   colnames(df_2allele), value = TRUE)[1]
drb15_col      <- grep("DRB1.15",   colnames(df_2allele), value = TRUE)[1]
core_col       <- grep("core",       colnames(df_long),    value = TRUE)[1]
percentile_col <- grep("percentile", colnames(df_long),    value = TRUE)[1]

cat("Datasets loaded:\n")
cat(" - df_wide:   ", nrow(df_wide),    "peptides x", ncol(df_wide),    "columns\n")
cat(" - df_2allele:", nrow(df_2allele), "peptides x", ncol(df_2allele), "columns\n")
cat(" - df_long:   ", nrow(df_long),    "rows x",     ncol(df_long),    "columns\n")
# ```
```
```{r alelle, echo=FALSE}
# ```{r tidy}
df_el <- df_wide %>%
  select(peptide, starts_with("binding.netmhciipan_el.percentile")) %>%
  pivot_longer(
    -peptide,
    names_to     = "allele",
    names_prefix = "binding.netmhciipan_el.percentile.",
    values_to    = "el_percentile"
  )
# ```
```

#Top binder Identification
```{r binder Identification, echo=FALSE}
top_binders <- df_wide %>%
  arrange(.data[[median_col]]) %>%
  slice_head(n = 20) %>%
  select(peptide, all_of(median_col))

top_binders %>%
  kable(
    caption   = "Top 20 peptides by global median EL percentile (lower = stronger)",
    col.names = c("Peptide", "Global Median EL Percentile"),
    digits    = 2
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(
    which(top_binders[[median_col]] <= 2),
    background = "#fde0dc", bold = TRUE
  ) %>%
  row_spec(
    which(top_binders[[median_col]] > 2 & top_binders[[median_col]] <= 10),
    background = "#fff3cd"
  )
```

# Binding Heatmap

```{r heatmap, fig.height=12, fig.width=14}
heat_mat <- df_wide %>%
  select(peptide, starts_with("binding.netmhciipan_el.percentile")) %>%
  column_to_rownames("peptide") %>%
  as.matrix()

colnames(heat_mat) <- gsub("binding.netmhciipan_el.percentile.", "", colnames(heat_mat))

pheatmap(
  heat_mat,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  color        = colorRampPalette(c("#2166ac", "white", "#d73027"))(100),
  breaks       = seq(0, 100, length.out = 101),
  fontsize_row = 6,
  fontsize_col = 7,
  main         = "MHC-II EL Binding Percentile Heatmap"
)
```

# Promiscuous Binder Selection

```{r promiscuous-table}
coverage <- df_el %>%
  filter(el_percentile <= 10) %>%
  group_by(peptide) %>%
  summarise(n_alleles_covered = n_distinct(allele)) %>%
  arrange(desc(n_alleles_covered))

coverage %>%
  slice_head(n = 10) %>%
  kable(
    caption   = "Top 10 promiscuous binders (EL <= 10%)",
    col.names = c("Peptide", "No. of Alleles Covered"),
    align     = "lr"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


```{r coverage}
ggplot(coverage,
       aes(x    = reorder(peptide, n_alleles_covered),
           y    = n_alleles_covered,
           fill = n_alleles_covered)) +
  geom_col() +
  scale_fill_gradient(low = "#c6dbef", high = "#2166ac") +
  coord_flip() +
  labs(
    title    = "Promiscuous MHC-II Binders (EL <= 10%)",
    subtitle = sprintf("%d peptides bind >= 1 allele at EL <= 10%%", nrow(coverage)),
    x        = "Peptide",
    y        = "Number of Alleles Covered",
    fill     = "Alleles"
  ) +
  theme_minimal(base_size = 10)
```
# Allele-Specific Analysis: DRB1*12:02 vs DRB1*15:02

```{r allele-classify}
df_2allele <- df_2allele %>%
  mutate(binder_class = case_when(
    .data[[global_col]] <= 10 ~ "Strong",
    .data[[global_col]] <= 30 ~ "Moderate",
    TRUE                      ~ "Weak"
  ))

n_strong   <- sum(df_2allele$binder_class == "Strong")
n_moderate <- sum(df_2allele$binder_class == "Moderate")
```

There are **`r n_strong`** strong binders and **`r n_moderate`** moderate binders
across both DRB1 alleles.

```{r allele-plot}
ggplot(df_2allele,
       aes(x     = .data[[drb12_col]],
           y     = .data[[drb15_col]],
           color = binder_class)) +
  geom_point(size = 2.5, alpha = 0.85) +
  geom_text_repel(
    data = df_2allele %>% filter(binder_class == "Strong"),
    aes(label = peptide), size = 2.5, box.padding = 0.4
  ) +
  scale_color_manual(
    values = c(Strong = "#d73027", Moderate = "#fc8d59", Weak = "grey70")
  ) +
  geom_hline(yintercept = 10, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 10, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  labs(
    title    = "DRB1*12:02 vs DRB1*15:02 - Binding Comparison",
    subtitle = "Labels shown for strong binders only (EL <= 10%)",
    x        = paste("EL Percentile -", drb12_col),
    y        = paste("EL Percentile -", drb15_col),
    color    = "Binder Class",
    caption  = "Dashed lines = 10th percentile threshold"
  ) +
  theme_minimal()
```

# Core Motif & Positional Analysis

```{r seqlogo, fig.height=3.5}
cores_top   <- df_long %>% filter(.data[[percentile_col]] <= 5) %>% pull(all_of(core_col))
n_cores     <- length(cores_top)
n_pep_logo  <- df_long %>% filter(.data[[percentile_col]] <= 5) %>% pull(peptide) %>% n_distinct()
n_alle_logo <- df_long %>% filter(.data[[percentile_col]] <= 5) %>% pull(allele)  %>% n_distinct()

ggseqlogo(cores_top, method = "bits") +
  annotate("rect",
           xmin = c(0.5, 3.5, 5.5, 8.5), xmax = c(1.5, 4.5, 6.5, 9.5),
           ymin = -Inf, ymax = Inf, fill = "#fdae61", alpha = 0.2) +
  annotate("text", x = c(1, 4, 6, 9), y = -0.1,
           label = "^", color = "#d73027", size = 3) +
  labs(
    title    = "MHC-II Core Binding Motif (EL <= 5th percentile)",
    subtitle = sprintf("%d cores | %d peptides | %d alleles | ^ = anchor P1/P4/P6/P9",
                       n_cores, n_pep_logo, n_alle_logo),
    x = "Core Position",
    y = "Information Content (bits)"
  )

```
 
```{r pos-dist}
pos_df <- df_long %>%
  filter(.data[[percentile_col]] <= 10) %>%
  distinct(peptide, start, end)

ggplot(pos_df, aes(x = start)) +
  geom_histogram(binwidth = 20, fill = "#4393c3", color = "white", alpha = 0.85) +
  geom_rug(aes(x = start), color = "#2166ac", alpha = 0.5) +
  geom_text(
    data  = pos_df %>% slice_min(start, n = 5, with_ties = FALSE),
    aes(x = start, y = 0, label = peptide),
    angle = 90, vjust = -0.3, hjust = 0, size = 2.5, color = "#d73027"
  ) +
  labs(
    title    = "Positional Distribution of Top Binders (EL <= 10%)",
    subtitle = sprintf("%d unique peptides", nrow(pos_df)),
    x        = "Start Position in Antigen (aa)",
    y        = "Count",
    caption  = "Rug = individual positions. Red labels = 5 N-terminal-most peptides."
  ) +
  theme_minimal(base_size = 11)
```

# Clustering by Binding Profile

```{r optimal-k, fig.width=10, fig.height=4}
mat_scaled <- scale(heat_mat)

p_elbow <- fviz_nbclust(mat_scaled, kmeans, method = "wss", k.max = 10) +
  geom_vline(xintercept = 4, linetype = "dashed", color = "#d73027") +
  labs(title = "Elbow Method", x = "k", y = "Within-Cluster SS")

p_sil <- fviz_nbclust(mat_scaled, kmeans, method = "silhouette", k.max = 10) +
  labs(title = "Silhouette Method", x = "k", y = "Avg Silhouette Width")

plot_grid(p_elbow, p_sil, ncol = 2)
```


```{r kmeans}
set.seed(42)
k_chosen        <- 4
km              <- kmeans(mat_scaled, centers = k_chosen, nstart = 25)
df_wide$cluster <- factor(km$cluster)

pca_res          <- prcomp(mat_scaled)
pca_df           <- as.data.frame(pca_res$x[, 1:2])
pca_df$peptide   <- rownames(mat_scaled)
pca_df$cluster   <- df_wide$cluster
pca_df$median_el <- df_wide[[median_col]]
var_exp          <- round(summary(pca_res)$importance[2, 1:2] * 100, 1)

label_df <- pca_df %>%
  group_by(cluster) %>%
  slice_min(median_el, n = 3, with_ties = FALSE)

ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(aes(size = -median_el), alpha = 0.8) +
  stat_ellipse(level = 0.85, linetype = "dashed") +
  geom_label_repel(
    data = label_df, aes(label = peptide),
    size = 2.5, box.padding = 0.4, show.legend = FALSE
  ) +
  scale_size_continuous(
    name   = "Binding strength",
    labels = function(x) paste0(-round(x, 1), "th pctile")
  ) +
  labs(
    title    = "Peptide Clusters by MHC-II Binding Profile (PCA)",
    subtitle = sprintf("k = %d | Labels = top 3 binders/cluster | Size proportional to binding strength",
                       k_chosen),
    x        = sprintf("PC1 (%s%% variance)", var_exp[1]),
    y        = sprintf("PC2 (%s%% variance)", var_exp[2]),
    color    = "Cluster",
    caption  = "Dashed ellipses = 85% confidence regions"
  ) +
  theme_minimal(base_size = 11)
```

```{r cluster-boxplot}
df_wide %>%
  ggplot(aes(x = cluster, y = .data[[median_col]], fill = cluster)) +
  geom_boxplot(outlier.shape = 21, width = 0.5, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.6) +
  geom_hline(yintercept = c(2, 10), linetype = "dashed",
             color = c("#d73027", "#fc8d59"), linewidth = 0.7) +
  annotate("text", x = 0.55, y = 2,
           label = "Strong (<=2%)",    hjust = 0, size = 3, color = "#d73027") +
  annotate("text", x = 0.55, y = 10,
           label = "Moderate (<=10%)", hjust = 0, size = 3, color = "#fc8d59") +
  labs(
    title    = "EL Percentile Distribution by Cluster",
    subtitle = "Lower = stronger MHC-II binder",
    x        = "Cluster",
    y        = "Global Median EL Percentile",
    caption  = "Dashed lines = strong (<=2%) and moderate (<=10%) thresholds"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

```{r cluster-table}
df_wide %>%
  group_by(cluster) %>%
  summarise(
    `Peptides (n)`     = n(),
    `Median EL`        = round(median(.data[[median_col]]), 2),
    `Min EL`           = round(min(.data[[median_col]]),    2),
    `Strong (<=2%)`    = sum(.data[[median_col]] <= 2),
    `Moderate (<=10%)` = sum(.data[[median_col]] <= 10),
    `Examples`         = paste(head(peptide, 2), collapse = ", ")
  ) %>%
  arrange(`Median EL`) %>%
  kable(caption = "Cluster summary - sorted by median EL percentile") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```


# Final Vaccine Candidate Shortlist
```{r shortlist}
final_candidates <- df_wide %>%
  filter(
    .data[[median_col]] <= 15,
    rowSums(select(., starts_with("binding.netmhciipan_el.percentile")) <= 10) >= 10
  ) %>%
  arrange(.data[[median_col]]) %>%
  select(peptide, all_of(median_col), cluster)

write_csv(final_candidates, "final_vaccine_candidates.csv")

final_candidates %>%
  kable(
    caption   = sprintf("%d final vaccine epitope candidates (EL <= 15%%, >= 10 alleles covered)",
                        nrow(final_candidates)),
    col.names = c("Peptide", "Global Median EL Percentile", "Cluster"),
    digits    = 2
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "bordered"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE, background = "#2166ac", color = "white")
```


# Session Info

```{r session-info}
sessionInfo()
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
